### 2023-12-21 sections/2-methods.tex
updated through the methods section

We perform some na\"{i}ve auto-tuning to select the optimal $k$-NN algorithm to use with a given dataset.
With this approach, we begin by taking a random sample of no more than 10\% of the data.
We build a cluster tree with this sample using Algorithm~\ref{alg:methods:clustering:partition}, and select all clusters at some user-specified depth in the tree, with a default depth of 10.
Using these clusters' centers as queries, and a user-specified value of $k$, we record the time taken for $k$-NN search on the sample using each of the four algorithms described in Section~\ref{subsec:methods:knn-search}.
We select the algorithm which had the lowest total search time over all the queries. 
Note that even though we select the optimal algorithm based on use with some user-specified value of $k$, we still allow search with any value of $k$.

@@

We expect CAKES to perform well as the cardinality of datasets increases, and to explore this scaling behavior, we augmented datasets in the ANN-benchmarks suite by adding random points within a 
small noise threshold of existing points; this technique has been used in a number of bioinformatics contexts~\cite{kumar2009augmented, kumar2010recognition, daniels2012smurflite, velasco2018combining, lu2020evolution}. This allowed us to isolate the effect of \emph{cardinality} on the performance of CAKES 
from that of other factors such as dimensionality, metric, or topological structure of the dataset.

To elaborate, we start with an original dataset $X$ from the ANN-benchmark suite, a user-specified noise tolerance $\epsilon$, and a mutliplier $m$.
For each datum $x \in X$, we create $m$ new data points within a distance $\epsilon$ of $x$. To construct a synthetic data point $x'$, we choose $d$ random numbers 
$r_1, r_2, \cdots , r_d$ in the interval $[-\tfrac{\epsilon}{\sqrt{d}}, \tfrac{\epsilon}{\sqrt{d}}]$ and add them componentwise to the original datapoint. 
Since $\sqrt{d \cdot (\tfrac{\epsilon}{\sqrt{d}})^2 } = \sqrt{d \cdot \tfrac{\epsilon^2}{d}}  = \sqrt{\epsilon^2} = \epsilon$, this ensures that any synthetic point is within a distance of $\epsilon$ from a real point in the original dataset.

>>>>

We perform some na\"{i}ve auto-tuning to select the optimal $k$-NN algorithm to use with a given dataset.
We start by taking the center of every cluster at a low depth, e.g. 10, in the cluster tree as a query.
This gives us a small set of queries which are representative of the dataset.
Using these clusters' centers as queries, and a user-specified value of $k$, we record the time taken for $k$-NN search on the sample using each of the three algorithms described in Section~\ref{subsec:methods:knn-search}.
We select the fastest over all the queries as the optimal algorithm for the dataset and value of $k$.
Note that even though we select the optimal algorithm based on use with some user-specified value of $k$, we still allow search with any value of $k$.

@@

Based on our asymptotic complexity analyses, we expect CAKES to perform well on datasets with low LFD, and for that performance to scale sub-linearly with the cardinality of the dataset.
To test this hypothesis, we use some datasets from the ANN-benchmarks suite~\cite{aumuller2020ann} and synthetically augment them to generate similar datasets with much higher cardinalities.
We do the same with a large random dataset of uniformly distributed points in a hypercube.
We then compare the performance of CAKES to that of other algorithms on the original datasets and the synthetically augmented datasets.

To elaborate on the augmentation process, we start with an original dataset from the ANN-benchmarks suite.
Let $X$ be the dataset, $d$ be its dimensionality, $\epsilon$ be a user-specified noise level and $m$ be a user-specified integer multiplier.
For each datum $x \in X$, we create $m$ new data points within a distance $\epsilon \cdot ||x||$ of $x$ where $||x||$ is the euclidean distance from $x$ to the origin.
We construct a random vector $r$ of $d$ dimensions in the hypersphere of radius $\epsilon$ centered at the origin.
We then add $r$ to $x$ to get a new point $x'$.
Since $||r|| = \epsilon$, we have that $||x - x'|| \leq \epsilon$, i.e. $x'$ is within a distance $\epsilon$ of $x$.
We repeat this process $m$ times for each $x \in X$ to get a new dataset $X'$.

This augmentation process preserves the topological structure of the original dataset, but increases its cardinality by a factor of $m$, allowing us to isolate the effect of cardinality on search performance from that of other factors such as dimensionality, choice of metric, or the topological structure of the dataset.
